Appendix 1.B: Analysing global trends in within-country inequality

### 2023-01-27 PhD_pages/data_appendices/1_A_prepare_PIP_and_WID.qmd
Further work on comparison app (allow any combo of variables)

df_pip<- df_pip %>%
  mutate(Bottom_50_share = decile1_share + decile2_share + decile3_share + decile4_share + decile5_share,
          P90_P10_ratio = decile9_thr/decile1_thr)

@@

#  Calculate P90_P10_ratio
df_wid<- df_wid %>%
    mutate(P90_P10_ratio = `P90...income.threshold.of.the.top.10.`/P10...income.threshold)

@@

```{r}
df_wid<- df_wid %>%
    rename("Gini" = "Gini.coefficient",
         "Top_10_share" = "P90.P100...share.of.the.top.10.",
         "Bottom_50_share"= "P0.P50...share.of.the.bottom.50.",
         "P90" = "P90...income.threshold.of.the.top.10.",
         "P10" = "P10...income.threshold") %>%
    select(
        Year,
        Entity,
        Gini,
        Top_10_share,
        # MLD,
        P90_P10_ratio,
        Bottom_50_share,
        P90,
        P10)

@@

@@

::: {.callout-note collapse="true" appearance="minimal"}
### Region classfications

>>>>

df_pip<- df_pip %>%
  mutate(Bottom_50_share = decile1_share + decile2_share + decile3_share + decile4_share + decile5_share,
          P90_P10_ratio = decile9_thr/decile1_thr,
          gini = gini * 100)

@@

#  Calculate P90_P10_ratio
df_wid<- df_wid %>%
    mutate(P90_P10_ratio = `P90...income.threshold.of.the.top.10.`/P10...income.threshold,
    Gini = Gini.coefficient * 100)

@@

```{r}
df_wid<- df_wid %>%
    rename("Top_10_share" = "P90.P100...share.of.the.top.10.",
         "Bottom_50_share"= "P0.P50...share.of.the.bottom.50.",
         "P90" = "P90...income.threshold.of.the.top.10.",
         "P10" = "P10...income.threshold") %>%
    select(
        Year,
        Entity,
        Gini,
        Top_10_share,
        # MLD,
        P90_P10_ratio,
        Bottom_50_share,
        P90,
        P10)

@@

## LIS data

from owid import catalog

# results = catalog.find('population')
# owid_pop = results[results['path'] == 'garden/owid/latest/key_indicators/population']

# df_wid<- read.csv("https://raw.githubusercontent.com/owid/notebooks/main/BetterDataDocs/PabloArriagada/WID/data/final/wid_pretax.csv")

# A local version, in case internet isn't available 
# df_wid<- read.csv("data/original/wid.csv")

@@

::: {.callout-note collapse="true" appearance="minimal"}
### Population

### 2023-01-27 PhD_pages/data_appendices/1_B_within_country_trends.qmd
Further work on comparison app (allow any combo of variables)

import pandas as pd
import numpy as np
from plotnine import *
import plotly.graph_objects as go
```

Load in the data
```{python}

df_wid = pd.read_csv("data/clean/wid.csv")
df_pip = pd.read_csv("data/clean/pip.csv")
df_regions = pd.read_csv("data/clean/WID_region_mapping.csv")
df_pop = pd.read_csv("data/clean/population.csv")

@@

@@

# df_wid = pd.read_csv("PhD_pages/data_appendices/data/clean/wid.csv")
# df_pip = pd.read_csv("PhD_pages/data_appendices/data/clean/pip.csv")
# df_regions = pd.read_csv("PhD_pages/data_appendices/data/clean/WID_region_mapping.csv")
# df_pop = pd.read_csv("PhD_pages/data_appendices/data/clean/population.csv")

A key difficulty of comparing trends is that the data is incomplete coverage. The paper uses two methods to calculate and compare aggregate within-country trends across incomplete data: using reference years and using year fixed effects in a regression.

## Method 1: Reference years

I need to allow for the income/consumption issue (currently the data includes both – i.e. multi observations per country year).

### Mapping data to a reference year
This function grabs the observation, by a grouping variable (e.g. by country), for which a reference variable (e.g. year) is closest to a reference value (e.g. 2002).
You can specify a maximum distance from the reference value, beyond which no match will be returned.
Because there may be tie-breaks – matches equally distant above or below the reference value – there is an argmuent to specify how these tie-breaks should resolved.
```{python}

@@

@@

I test this function here:
```{python}

test = closest_to_reference(
  df = df_wid, 
  reference_val = 2002,
  max_dist_from_ref = 5,
  reference_col = 'Year',
  group_by_col = 'Entity',
  value_col = 'Gini',
  tie_break = 'below'
  )

test.head()

### Obtaining a pair of observations

This function runs the `closest_to_reference` function twice, over a pair of reference values.
Tie-breaks are settled so as to maximise the gap between the two reference values.
You can also specify a minimum distance between the two observations (e.g. pairs of matches that fall less than X years apart will be dropped).

```{python}
# Merge matches for different reference points
def merge_two_ref_matches(df, reference_vals, max_dist_from_refs, min_dist_between, reference_col, group_by_col, value_col):

@@

  # Find matches for lower reference value
  lower_ref_matches = closest_to_reference(df, reference_vals[0], max_dist_from_refs[0], reference_col, group_by_col, value_col, 'below')

  # Find matches for higher reference value
  higher_ref_matches = closest_to_reference(df, reference_vals[1], max_dist_from_refs[1], reference_col, group_by_col, value_col, 'above')

@@

I test this function here:
```{python}
test = merge_two_ref_matches(
  df = df_wid, 
  reference_vals = [2000, 2010], 
  max_dist_from_refs = [5, 4],
  min_dist_between = 9,
  reference_col = 'Year',
  group_by_col = 'Entity',
  value_col = 'Gini'
  )

test.head()

#### Allowing for the different welfare concepts in the PIP data

As noted in .... (add backlink) ... the PIP data contains both income and consumption observations.
This function produces matched pairs of observations from that data in which only one pair is selected for each value of the grouping variable (i.e. each country).
Priority is given to pairs for which both observations relate to income. Second priority is given to pairs where both observations relate to consumption.
Only if neither pair is available then it will return a mixed pair of observations if that is available.
```{python}
# For PIP run this three times – first filtering data for just consumpion only, then with income only, then with a dataset that prefers income over consumption
def pip_welfare_routine(df, reference_vals, max_dist_from_refs, min_dist_between, reference_col, group_by_col, value_col):

@@

I test this function here:
```{python}

test = pip_welfare_routine(

  df = df_pip,
  reference_vals = [1986, 2016], 
  max_dist_from_refs = [5, 5], 
  min_dist_between = 30, 
  reference_col = 'Year',
  group_by_col = 'Entity',
  value_col = 'Gini'

  )

test.head()
```

### Merging reference year aligned data from WID and PIP datasets

@@

def prep_wid_pip_data(reference_vals, max_dist_from_refs, min_dist_between, reference_col, group_by_col, value_col):
  pip_matches = pip_welfare_routine(
    df = df_pip,
    reference_vals = reference_vals,
    max_dist_from_refs = max_dist_from_refs,
    min_dist_between = min_dist_between,
    reference_col = reference_col,
    group_by_col = group_by_col,
    value_col = value_col
  )

@@

  return ref_pairs
```

## Prepare plot and tables

```{python}
#| echo: false

@@

  # Make nicely formatted table
  plot_data_formatted = plot_data.copy()

  plot_data_formatted[x_axis] = plot_data_formatted[x_axis].map('{:.2f}'.format)
  plot_data_formatted[y_axis] = plot_data_formatted[y_axis].map('{:.2f}'.format)

  plot_data_table = go.Figure(data=[go.Table(
    header=dict(values=list(plot_data_formatted.columns)),
    cells=dict(values=plot_data_formatted.transpose().values.tolist() ))
    ])

@@

  plot = (ggplot(plot_data
  , aes(x_axis, y_axis, alpha=0.5))
  + geom_point(aes(color=region_col, size='avg_pop', alpha = 0.6))
  + facet_wrap('~ source')
  + geom_polygon(aes(x='x', y='y', alpha = 0.2), data=shaded_coord)
  + theme_light())

@@

  # ----- Return outputs ------
  plot_and_tables = {
    "table_data": plot_data,
    "table": plot_data_table,
    "plot": plot,
    "summary": df_summary
  }

@@

```{python}
pd.set_option('display.max_rows', None)

## 1990-2017
```{python}
# Specifications
reference_vals = [1990, 2017]
max_dist_from_refs = [4, 4]
min_dist_between = 23

# These parameters are unlikely to change:
reference_col = 'Year'
group_by_col = 'Entity'

::: {.panel-tabset}
## Gini

```{python}
# | echo: false
# Specifications

value_col = 'Gini'
# 'Gini'
# 'Top_10_share'
# 'Bottom_50_share'
# 'MLD'
# 'P90_P10_ratio' – not that helpful due to accuracy of low incomes

# Set the value which constitutes a significant change in the value being measured
  # E.g. for Gini tolerance = 0.02 (2 points)

tolerance = 0.02
outlier_cut_off_upper = None

```
::: {.panel-tabset}
## All countries
```{python}
# | echo: false

# Data from both sources is needed?
  # If this = 1 then a country will be incldued even if it is only availalbe from one source.
  # If = 2, then only countries with data for both sources will be included
source_count_requirement = 1

plot_and_tables = prep_plot_and_tables(
  reference_vals = reference_vals, 
  max_dist_from_refs = max_dist_from_refs,
  min_dist_between = min_dist_between,
  reference_col = reference_col,
  group_by_col = group_by_col,
  value_col = value_col,
  tolerance = tolerance,
  outlier_cut_off_upper = outlier_cut_off_upper,
  source_count_requirement = source_count_requirement
  )

```
::: {.panel-tabset}
## Plot
```{python}
plot_and_tables['plot']
```

## Summary
```{python}
plot_and_tables['summary']
```

## Data
```{python}

plot_data_results = plot_and_tables['table_data']
plot_data_results
```
:::

## Countries with data for both sources

```{python}
# | echo: false

# Data from both sources is needed?
  # If this = 1 then a country will be incldued even if it is only availalbe from one source.
  # If = 2, then only countries with data for both sources will be included
source_count_requirement = 2

plot_and_tables = prep_plot_and_tables(
  reference_vals = reference_vals, 
  max_dist_from_refs = max_dist_from_refs,
  min_dist_between = min_dist_between,
  reference_col = reference_col,
  group_by_col = group_by_col,
  value_col = value_col,
  tolerance = tolerance,
  outlier_cut_off_upper = outlier_cut_off_upper,
  source_count_requirement = source_count_requirement
  )

```
::: {.panel-tabset}
## Plot
```{python}
plot_and_tables['plot']
```

## Summary
```{python}
plot_and_tables['summary']
```

## Data
```{python}

plot_data_results = plot_and_tables['table_data']
plot_data_results
```
:::
:::

## Top 10% share

```{python}
# | echo: false
# Specifications

value_col = 'Top_10_share'

# Set the value which constitutes a significant change in the value being measured
  # E.g. for Gini tolerance = 0.02 (2 points)

```
::: {.panel-tabset}
## All countries
```{python}
# | echo: false

# Data from both sources is needed?
  # If this = 1 then a country will be incldued even if it is only availalbe from one source.
  # If = 2, then only countries with data for both sources will be included
source_count_requirement = 1

plot_and_tables = prep_plot_and_tables(
  reference_vals = reference_vals, 
  max_dist_from_refs = max_dist_from_refs,
  min_dist_between = min_dist_between,
  reference_col = reference_col,
  group_by_col = group_by_col,
  value_col = value_col,
  tolerance = tolerance,
  outlier_cut_off_upper = outlier_cut_off_upper,
  source_count_requirement = source_count_requirement
  )

```
::: {.panel-tabset}
## Plot
```{python}
plot_and_tables['plot']
```

## Summary
```{python}
plot_and_tables['summary']
```

## Data
```{python}

plot_data_results = plot_and_tables['table_data']
plot_data_results
```
:::

## Countries with data for both sources

```{python}
# | echo: false

# Data from both sources is needed?
  # If this = 1 then a country will be incldued even if it is only availalbe from one source.
  # If = 2, then only countries with data for both sources will be included
source_count_requirement = 2

plot_and_tables = prep_plot_and_tables(
  reference_vals = reference_vals, 
  max_dist_from_refs = max_dist_from_refs,
  min_dist_between = min_dist_between,
  reference_col = reference_col,
  group_by_col = group_by_col,
  value_col = value_col,
  tolerance = tolerance,
  outlier_cut_off_upper = outlier_cut_off_upper,
  source_count_requirement = source_count_requirement
  )

```
::: {.panel-tabset}
## Plot
```{python}
plot_and_tables['plot']
```

## Summary
```{python}
plot_and_tables['summary']
```

## Data
```{python}

plot_data_results = plot_and_tables['table_data']
plot_data_results
```
:::

:::

Next measure

:::

## Method 2: regression analysis

## Explore the date

from shiny import *
from plotnine import *
from pyodide.http import open_url
import pandas as pd

app_ui = ui.page_fluid(
    ui.output_plot("example_plot"),
)

        url = 'https://raw.githubusercontent.com/owid/notebooks/main/BetterDataDocs/JoeHasell/PIP/data/ppp_2017/final/PIP_data_public_download/full_dataset/cons_only/poverty_cons_only.csv'

        df = pd.read_csv(open_url(url))

        plot = (ggplot(df, aes('Year', 'headcount_ratio_365', color = 'Entity'))
        + geom_line())

        # d = {'col1': [1, 2], 'col2': [3, 4]}
        # df = pd.DataFrame(data=d)
        # plot = (ggplot(df, aes('col1', 'col2'))
        # + geom_point())

>>>>

from plotnine import *
import pandas as pd
import numpy as np

@@

This first block of code.....

@@

#  ---------------------------------
#  ---------- SECTION 1: Data prep functions -------------
#  ----------------------------------

@@

  df = df\
    .rename(columns={value_col: "value"})

@@

# Merge matches for different reference points
def merge_two_ref_matches(df, reference_vals, max_dist_from_refs, min_dist_between, reference_col, group_by_col, value_col):

@@

# Find matches for lower reference value
  lower_ref_matches = closest_to_reference(df, reference_vals[0], max_dist_from_refs, reference_col, group_by_col, value_col, 'below')

# Find matches for higher reference value
  higher_ref_matches = closest_to_reference(df, reference_vals[1], max_dist_from_refs, reference_col, group_by_col, value_col, 'above')

@@

  # For PIP run this three times – first filtering data for just consumpion only, then with income only, then with a dataset that prefers income over consumption
def pip_welfare_routine(df, reference_vals, max_dist_from_refs, min_dist_between, reference_col, group_by_col, value_col):

@@

```{python}
#  ---------------------------------
#  ---------- SECTION 2: Specifications -------------
#  ----------------------------------

# These parameters are not controlled in the app.
reference_col = 'Year'
group_by_col = 'Entity'

source_count_requirement = "all"
# source_count_requirement = "any"

reference_vals = [1990,2018]
max_dist_from_refs = 4
min_dist_between = 5

data = {"PIP": df_pip,
        "WID": df_wid}

data_to_match = {
  "sources": ["PIP", "WID", "PIP"],
  "var_names": ["Gini", "Top_10_share", "Top_10_share"]
  }

This second block of code we read in the data (this needs to be done in a different way in )

@@

#  ---------------------------------
#  ---------- SECTION 3: Read in data -------------
#  ----------------------------------

url = 'https://raw.githubusercontent.com/JoeHasell/personal_site/main/PhD_pages/data_appendices/data/clean/pip.csv'

df_pip = pd.read_csv(url)

url = 'https://raw.githubusercontent.com/JoeHasell/personal_site/main/PhD_pages/data_appendices/data/clean/wid.csv'

df_wid = pd.read_csv(url)

url = 'https://raw.githubusercontent.com/JoeHasell/personal_site/main/PhD_pages/data_appendices/data/clean/WID_region_mapping.csv'

df_regions = pd.read_csv(url)

url = 'https://raw.githubusercontent.com/JoeHasell/personal_site/main/PhD_pages/data_appendices/data/clean/population.csv'

df_pop = pd.read_csv(url)

#  ---------------------------------
#  ---------- SECTION 4: Run data functions and produce plot -------------
#  ----------------------------------

# Combine the source and varname strings element-wise to produce keys that will be used for the concat later
source_var_keys = [i + j for i, j in zip(data_to_match["sources"], data_to_match["var_names"])]

def prep_and_merge_multiple_vars(data_to_match, reference_vals, max_dist_from_refs, min_dist_between, reference_col, group_by_col):

  matches = []

  for i in range(len(data_to_match['sources'])):

    source = data_to_match["sources"][i]
    var = data_to_match["var_names"][i]
    print(i)
    print(source)
    print(var)
    df = data[source]

    if source == "PIP":

      matched = pip_welfare_routine(
          df = df,
          reference_vals = reference_vals,
          max_dist_from_refs = max_dist_from_refs,
          min_dist_between = min_dist_between,
          reference_col = reference_col,
          group_by_col = group_by_col,
          value_col = var
        )

    else:

      matched = merge_two_ref_matches(
          df = df,
          reference_vals = reference_vals,
          max_dist_from_refs = max_dist_from_refs,
          min_dist_between = min_dist_between,
          reference_col = reference_col,
          group_by_col = group_by_col,
          value_col = var
        )

    matches.append(matched)

  ref_matches = pd.concat(matches, keys=source_var_keys)

  # Tidy up indexes
  ref_matches = ref_matches.reset_index()

  ref_matches = ref_matches.drop('level_1', axis=1)

  ref_matches = ref_matches\
    .rename(columns={"level_0": "source_var"})

  return ref_matches

def prep_plot_and_tables(
  data_to_match,
  reference_vals, 
  max_dist_from_refs,
  min_dist_between,
  reference_col,
  group_by_col,
  tolerance,
  outlier_cut_off_upper,
  source_count_requirement):

  plot_data = prep_and_merge_multiple_vars(
    data_to_match = data_to_match,
    reference_vals = reference_vals, 
    max_dist_from_refs = max_dist_from_refs,
    min_dist_between = min_dist_between,
    reference_col = reference_col,
    group_by_col = group_by_col
  )

  # Store the names of the columns to be used onthe X and Y axis
  x_axis = f'value{reference_vals[0]}'
  y_axis = f'value{reference_vals[1]}'

  # Add a count by country – counting whether data is available from each source or not
  plot_data['source_count'] = plot_data.groupby(group_by_col)['source_var'].transform('count')

  # Apply source requirement (whether to include only observations with data from all sources)
  if source_count_requirement == "all":
    plot_data = plot_data.loc[plot_data['source_count'] == plot_data['source_count'].max()]

  plot = (ggplot(plot_data
  , aes(x_axis, y_axis, alpha=0.5))
  + geom_point(aes(color=region_col, size='avg_pop', alpha = 0.6))
  + facet_wrap('~ source_var')
  + geom_polygon(aes(x='x', y='y', alpha = 0.2), data=shaded_coord)
  + theme_light()
  + theme(legend_position='none')
  )

  group_yes_vars = ['source_var', region_col]
  group_no_vars = ['source_var']

    agg_level['global_pop_weights'] = agg_level.groupby('source_var')['avg_pop'].transform(lambda x: x/x.sum())

  # ----- Return outputs ------
  plot_and_tables = {
    "table_data": plot_data,
    "plot": plot,
    "summary": df_summary
  }

```{python}
plot_and_tables = prep_plot_and_tables(
    data_to_match = data_to_match,
    reference_vals = reference_vals, 
    max_dist_from_refs = max_dist_from_refs,
    min_dist_between = min_dist_between,
    reference_col = reference_col,
    group_by_col = group_by_col,
    tolerance = tolerance,
    outlier_cut_off_upper = outlier_cut_off_upper,
    source_count_requirement = source_count_requirement
    )

plot = plot_and_tables['plot']
summary = plot_and_tables['summary']

```{python}
plot

```{python}
summary

## Explore the data

from shiny import *
from plotnine import *
from pyodide.http import open_url
import pandas as pd
import numpy as np

#  ---------------------------------
#  ---------- Data prep functions -------------
#  ----------------------------------

# Merge matches for different reference points
def merge_two_ref_matches(df, reference_vals, max_dist_from_refs, min_dist_between, reference_col, group_by_col, value_col):

# Find matches for lower reference value
  lower_ref_matches = closest_to_reference(df, reference_vals[0], max_dist_from_refs, reference_col, group_by_col, value_col, 'below')

# Find matches for higher reference value
  higher_ref_matches = closest_to_reference(df, reference_vals[1], max_dist_from_refs, reference_col, group_by_col, value_col, 'above')

  # For PIP run this three times – first filtering data for just consumpion only, then with income only, then with a dataset that prefers income over consumption
def pip_welfare_routine(df, reference_vals, max_dist_from_refs, min_dist_between, reference_col, group_by_col, value_col):

#  ---------------------------------
#  ---------- Specifications -------------
#  ----------------------------------

# These parameters are not controlled in the app.
reference_col = 'Year'
group_by_col = 'Entity'

source_count_requirement = 2

#  ---------------------------------
#  ---------- App code -------------
#  ----------------------------------

def panel_box(*args, **kwargs):
    return ui.div(
        ui.div(*args, class_="card-body"),
        **kwargs,
        class_="card mb-3",
    )

app_ui = ui.page_fluid(
    {"class": "p-4"},
    ui.row(
        ui.column(
            4,
            panel_box(
                ui.input_slider("reference_vals", "Reference years", 1975, 2020, value=[1990,2018]),
                ui.input_numeric(
                            "max_dist_from_refs",
                            "=/- (years)",
                            4,
                        ),
                ui.input_numeric(
                            "min_dist_between",
                            "Min distance between observations (years)",
                            23,
                        ),
                ui.input_action_button(
                    "download", "Download data and image", class_="btn-primary w-100"
                ),
            ),
            ui.navset_tab_card(
                ui.nav(
                    "Compare sources",
                    ui.input_select(
                "metric",
                "Metric",
                {
                    "Gini": "Gini",
                    "Top_10_share": "Top 10p.c. share",
                },
            ),
            ),
                ui.nav(
                  "Compare metrics"),
            ),
        ),
        ui.column(
            8,
            ui.output_plot("example_plot")
            ),
    ),
)

      url = 'https://raw.githubusercontent.com/JoeHasell/personal_site/main/PhD_pages/data_appendices/data/clean/pip.csv'

      df_pip = pd.read_csv(open_url(url))

      url = 'https://raw.githubusercontent.com/JoeHasell/personal_site/main/PhD_pages/data_appendices/data/clean/wid.csv'

      df_wid = pd.read_csv(open_url(url))

      url = 'https://raw.githubusercontent.com/JoeHasell/personal_site/main/PhD_pages/data_appendices/data/clean/WID_region_mapping.csv'

      df_regions = pd.read_csv(open_url(url))

      url = 'https://raw.githubusercontent.com/JoeHasell/personal_site/main/PhD_pages/data_appendices/data/clean/population.csv'

      df_pop = pd.read_csv(open_url(url))

      def prep_wid_pip_data(reference_vals, max_dist_from_refs, min_dist_between, reference_col, group_by_col, value_col):

        pip_matches = pip_welfare_routine(
          df = df_pip,
          reference_vals = reference_vals,
          max_dist_from_refs = max_dist_from_refs,
          min_dist_between = min_dist_between,
          reference_col = reference_col,
          group_by_col = group_by_col,
          value_col = value_col
        )

@@

        return ref_pairs

@@

@@

        plot = (ggplot(plot_data
        , aes(x_axis, y_axis, alpha=0.5))
        + geom_point(aes(color=region_col, size='avg_pop', alpha = 0.6))
        + facet_wrap('~ source')
        + geom_polygon(aes(x='x', y='y', alpha = 0.2), data=shaded_coord)
        + theme_light()
        + theme(legend_position='none')
        )

@@

        # ----- Return outputs ------
        plot_and_tables = {
          "table_data": plot_data,
          "plot": plot,
          "summary": df_summary
        }

@@

      plot_and_tables = prep_plot_and_tables(
          reference_vals = list(input.reference_vals()), 
          max_dist_from_refs = input.max_dist_from_refs(),
          min_dist_between = input.min_dist_between(),
          reference_col = reference_col,
          group_by_col = group_by_col,
          value_col = input.metric(),
          tolerance = tolerance,
          outlier_cut_off_upper = outlier_cut_off_upper,
          source_count_requirement = source_count_requirement
          )

      plot = plot_and_tables['plot']
